---
title: "Analyzing Player Metrics and Market Value in European Football"
author: "Abigail Chen, Maxwell Gerhart, Sanjana Menon"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Football is the most popular sport in the world, where major tournaments
like the FIFA World Cup and major leagues such as the English Premier
League, Ligue 1, Serie A, La Liga and the Bundesliga attract billions of
followers and viewers. Better known as soccer in the United States,
football is a simple game where two teams compete to score goals by
moving the ball into the opposing net, primarily using their feet.
Contrasting the simplicity of the game is the dynamics behind the
scenes, where clubs invest heavily in players after analyzing metrics
like market value, match statistics, and positional roles. This leads us
to our project, which examines how player attributes such as age,
minutes played, position, and goal contributions as well as the team
attribute of winning percentages relate to market valuation and ultimate
team success across the top 5 leagues, particularly for the 2024-2025
season.

# Research Questions

Our goal is to get a better understanding of how player and eventually,
team performance metrics relate to both market valuation and team
outcomes. In order to accomplish that, we ask: How does a player's age
and position influence their minutes played and market value? Do players
with higher goal contributions consistently have greater market worth?
How do team averages such as player age, minutes played, and goal
contributions correlate with win percentages across competitions?

Through visual analyses, including scatterplots, boxplot, heatmap, and
summary tables, we aim to understand how measurable performance
indicators influence both individual worth and collective outcomes in
the sport.

# Provenance of Our Data

For this project, we used four datasets to gather core performance
indicators in football, all obtained from Transfermarkt's Football Data
posted on Kaggle. Transfermarkt is a widely used football database that
tracks player market values, match statistics, transfers, and club
performance across global competitions. The dataset was created and
maintained by David Cariboo, who used a web-scraping tool called
transfermarkt-scraper along with Python scripts and SQL queries to
extract, clean, and publish the data. The full data pipeline and source
code are available on his GitHub. The purpose of his work is to keep
these datasets up to date and publicly available on well-known data
catalogs. The data is publicly available under the CC0 (Public Domain)
license, with an expected update frequency of once per week. Our primary
data source was Player Valuation while our secondary sources were Player
Appearances, Player Statistics & Demographics and Game Statistics.

## Primary Dataset

-   **Player Valuation**: This was our main dataset that provided market
    value data for players over time. Key attributes were used
    consistently throughout our visualizations, which were player_id,
    date, and market_value_in_eur. The case is a single player.

## Secondary Datasets

-   **Player Appearances**: This dataset includes detailed match-level
    data on individual players. We used attributes like game_id,
    player_id, player_club_id, goals, assists, and minutes_played to
    analyze player contributions across matches. The case is a single
    player.

-   **Player Statistics & Demographics**: This dataset includes personal
    and positional data for players. Our analysis focused on position,
    while also referencing identifiers like player_id and broader
    details such as date_of_birth for age related analysis. The case is
    a single player.

-   **Game Statistics**: This dataset includes match-level team
    performance indicators. We used game_id, home_club_id, away_club_id,
    home_club_goals, and away_club_goals to calculate our new attribute,
    win percentages. Other fields, including competition_id, season, and
    round, offer useful structural context for game outcomes. The case
    is a single game.

# Implementation of FAIR & CARE Principles

## FAIR Principles

- **Findable**:  
  The dataset is hosted on Kaggle, a widely-used platform that assigns a permanent identifier and includes metadata such as title, description, and file previews. These features ensure the dataset is easily located through web search and academic repositories.

- **Accessible**:  
  Access is open with a Kaggle account, and the data can be downloaded in CSV format without technical barriers. This accessibility supports broad use across educational and research communities.

- **Interoperable**:  
  Provided in a standard tabular structure (CSV), the dataset can be integrated into common data science workflows using tools like R, Python, and SQL. Column headers and values follow widely understood naming conventions, which supports semantic clarity.

- **Reusable**:  
  The dataset includes basic metadata, source attribution, and a clear license (CC0: Public Domain Dedication), allowing for reuse without restriction. Users can confidently integrate and repurpose the data for academic, commercial, or exploratory purposes.

## CARE Principles

- **Collective Benefit**:  
  The dataset supports public interest research in sports analytics, education, and predictive modeling. By promoting open access to structured player data, it fosters innovation and insight in football analysis.

- **Authority to Control**:  
  As the dataset aggregates public sports data without identifiable private or community information, issues of consent and governance are minimal. However, transparency about the original data sources could further strengthen accountability.

- **Responsibility**:  
  While the dataset itself poses low ethical risk, users must remain mindful of how derived insights are applied—particularly in contexts like player valuation or contract negotiation, where misuse could influence careers or reinforce bias.

- **Ethics**:  
  The dataset avoids sensitive or protected attributes and reflects common practices in sports statistics. Ethical use demands acknowledgment of context and purpose, ensuring analyses do not misrepresent players or their performance.

# Exploratory Data Analysis (EDA)

### Age, Playing Time & Position

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
# Load required libraries
library(readr)
library(ggplot2)
library(dplyr)

# Load data
appearances <- read_csv("C:/Users/maxwe/OneDrive - The Pennsylvania State University/Downloads/data/appearances.csv")

players <- read_csv("C:/Users/maxwe/OneDrive - The Pennsylvania State University/Downloads/data/players.csv")

player_valuations <- read_csv("C:/Users/maxwe/OneDrive - The Pennsylvania State University/Downloads/data/player_valuations.csv")

games <- read_csv("C:/Users/maxwe/OneDrive - The Pennsylvania State University/Downloads/data/games.csv")

clubs <- read_csv("C:/Users/maxwe/OneDrive - The Pennsylvania State University/Downloads/data/clubs.csv")

# Define top 5 European football leagues
top5_leagues <- c("GB1", "FR1", "IT1", "ES1", "L1")  # Premier League, Ligue 1, Serie A, La Liga, Bundesliga

# Define season date range
season_start <- as.Date("2024-08-01")
season_end <- as.Date("2025-06-30")

# Clean and filter appearances data
appearances_top5 <- appearances %>%
  filter(competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Clean and filter player valuations data
player_valuations_top5 <- player_valuations %>%
  filter(player_club_domestic_competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Prepare player data with standardized birth date
players <- players %>%
  mutate(birth_date = as.Date(substr(date_of_birth, 1, 10)))  # Extract date part

# Filter valuations for current season and calculate player age
valuations_season <- player_valuations_top5 %>%
  filter(between(date, season_start, season_end)) %>%
  left_join(players %>% select(player_id, birth_date), by = "player_id") %>%
  mutate(age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25)

# Calculate minutes played per player for current season
player_season_stats <- appearances_top5 %>%
  filter(between(date, season_start, season_end)) %>%
  group_by(player_id) %>%
  summarize(
    total_minutes = sum(minutes_played, na.rm = TRUE),
    matches_played = n(),
    avg_minutes_per_match = total_minutes / matches_played,
    .groups = "drop"
  )

# Merge most recent valuation, minutes played, and player position
player_analysis <- valuations_season %>%
  group_by(player_id) %>%
  arrange(desc(date)) %>%
  slice(1) %>%
  ungroup() %>%
  left_join(player_season_stats, by = "player_id") %>%
  left_join(players %>% select(player_id, position), by = "player_id")

# Create age bins and rounded age for analysis
player_analysis <- player_analysis %>%
  mutate(
    age_bin = cut(age, breaks = seq(16, 40, by = 2), include.lowest = TRUE),
    age_rounded = floor(age)
  )

# Create the scatter plot with facets for each position
ggplot(player_analysis, aes(x = age, y = total_minutes)) +
  geom_point(aes(color = position), alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, aes(color = position), alpha = 0.2) +
  facet_wrap(~ position, ncol = 2) +
  scale_color_manual(values = c(
    "Attack" = "#FF5733", 
    "Midfield" = "#33A1FD", 
    "Defender" = "#9370DB", 
    "Goalkeeper" = "#2ECC71"
  )) +
  scale_x_continuous(breaks = seq(16, 40, by = 4), limits = c(16, 40)) +
  labs(
    title = "Age vs Minutes Played by Position",
    subtitle = "Top 5 European Leagues (2024–2025 Season)",
    x = "Age (years)",
    y = "Total Minutes Played",
    color = "Position"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "lightgray", color = NA),
    strip.text = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

Figure 1: This scatter plot displays the relationship between player age
and total minutes played during the 2024–2025 season across the top 5
European leagues. Faceted by position, it highlights positional
differences in playing time patterns by age group, with trend lines
indicating general age-performance curves.

### Insight

This visual allows us to analyze the peak age based on minutes played for each position. From the scatter plot, we can see that attackers tend to peak earliest, around ages 24 to 26, while goalkeepers peak later, around ages 29 to 30. Midfielders and defenders are more evenly spread, peaking between 25 and 28.

It's important to keep this in mind when comparing market value across different age groups and positions.

### Age, Position & Market Value

```{r, echo=FALSE, fig.width=12, fig.height=6}
# Load required libraries
library(ggplot2)
library(dplyr)

# Define top 5 European football leagues
top5_leagues <- c("GB1", "FR1", "IT1", "ES1", "L1")

# Define season date range
season_start <- as.Date("2024-08-01")
season_end <- as.Date("2025-06-30")

# Clean and filter appearances data
appearances_top5 <- appearances %>%
  filter(competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Clean and filter player valuations data
player_valuations_top5 <- player_valuations %>%
  filter(player_club_domestic_competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Prepare player data with standardized birth_date
players_clean <- players %>%
  mutate(birth_date = as.Date(substr(date_of_birth, 1, 10))) %>%
  select(player_id, birth_date, position)

# Filter valuations for current season and join birth date
valuations_season <- player_valuations_top5 %>%
  filter(between(date, season_start, season_end)) %>%
  left_join(players_clean %>% select(player_id, birth_date), by = "player_id") %>%
  mutate(age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25)

# Get most recent valuation per player and add position
player_analysis <- valuations_season %>%
  group_by(player_id) %>%
  arrange(desc(date)) %>%
  slice(1) %>%
  ungroup() %>%
  left_join(players_clean %>% select(player_id, position), by = "player_id")

# Create age bins
player_analysis <- player_analysis %>%
  mutate(
    age_bin = factor(case_when(
      age < 19 ~ "16-18",
      age < 21 ~ "19-20",
      age < 23 ~ "21-22",
      age < 25 ~ "23-24",
      age < 27 ~ "25-26",
      age < 29 ~ "27-28",
      age < 31 ~ "29-30",
      age < 33 ~ "31-32",
      age < 35 ~ "33-34",
      age < 37 ~ "35-36",
      age < 39 ~ "37-38",
      age <= 40 ~ "39-40"
    ), levels = c("16-18", "19-20", "21-22", "23-24", "25-26", "27-28", 
                  "29-30", "31-32", "33-34", "35-36", "37-38", "39-40")),
    position = ifelse(is.na(position), "Unknown", position)
  )

# Summarize for heatmap
heatmap_data <- player_analysis %>%
  filter(!is.na(market_value_in_eur) & !is.na(position) & !is.na(age_bin)) %>%
  group_by(age_bin, position) %>%
  summarize(
    avg_market_value = mean(market_value_in_eur, na.rm = TRUE),
    median_market_value = median(market_value_in_eur, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    avg_market_value_millions = avg_market_value / 1e6,
    median_market_value_millions = median_market_value / 1e6
  )

# Ensure all age-position combinations are included
all_combinations <- expand.grid(
  age_bin = levels(player_analysis$age_bin),
  position = unique(player_analysis$position)
)

# Join with full combinations and fill missing with NA
heatmap_data <- all_combinations %>%
  left_join(heatmap_data, by = c("age_bin", "position")) %>%
  mutate(
    avg_market_value_millions = ifelse(is.na(avg_market_value_millions), NA, avg_market_value_millions),
    label = ifelse(is.na(avg_market_value_millions), "N/A", sprintf("%.0f€", avg_market_value_millions))
  )

# Order positions
heatmap_data$position <- factor(
  heatmap_data$position, 
  levels = c("Goalkeeper", "Defender", "Midfield", "Attack", "Unknown")
)

# Plot the heatmap
ggplot(heatmap_data, aes(x = age_bin, y = position, fill = avg_market_value_millions)) +
  geom_tile() +
  geom_text(aes(label = label), color = "white", size = 5, fontface = "bold") +
  scale_fill_gradient(
    low = "blue", high = "orange", 
    na.value = "grey50",
    name = "Avg. Market Value (M€)",
    breaks = seq(0, ceiling(max(heatmap_data$avg_market_value_millions, na.rm = TRUE)), by = 10)
  ) +
  labs(
    title = "Age vs Position Heatmap",
    subtitle = "(Top 5 European Leagues, 2024–2025 Season)",
    x = "Age Group",
    y = "Position",
    caption = "Data includes most recent valuation per player"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(face = "bold", size = 12),
    legend.position = "right",
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )

```

Figure 2: This heatmap visualizes average market value (in millions of euros) across different age groups and positions. Warmer colors represent higher values, with labels showing rounded figures. This allows easy comparison of how age and role impact player valuation in elite European football.

### Insight

We see that players tend to be more valuable in positions higher up the pitch. Additionally, there is a clear peak in market value just before the typical peak in minutes played. This suggests that clubs invest more in players approaching their prime, anticipating strong performance and longevity. The alignment between expected trends and the data reinforces the reliability of this insight.

### Playing Time, Market Value & Position

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
# Load required libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyverse)
library(scales)

# Aggregate player appearances
player_stats <- appearances_top5 %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_goals = sum(goals, na.rm = TRUE),
    total_appearances = n(),
    total_minutes = sum(minutes_played, na.rm = TRUE),
    avg_minutes_per_appearance = total_minutes / total_appearances,
    .groups = "drop"
  )

# Average market value
player_values <- valuations_season %>%
  group_by(player_id) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE), .groups = "drop")

# Merge stats and values
player_analysis <- player_stats %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(
    players %>% select(player_id, position),
    by = "player_id"
  )

# Plot: Minutes vs. Market Value
appearance_plot <- ggplot(
  player_analysis,
  aes(x = total_minutes, y = avg_market_value, color = position)
  ) +
  geom_point(alpha = 0.5, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Market Value vs. Minutes Played",
    subtitle = "Top 5 European Leagues (2024–2025 Season)",
    x = "Total Minutes Played",
    y = "Avg. Market Value (EUR)",
    color = "Position"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 30, 10)) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

print(appearance_plot)

```
Figure 3: This scatterplot displays the relationship between total minutes played and average market value (in millions of Euros) for players in the top 5 European leagues during the 2024-2025 season. Each point represents a player, colored by their position: Attack (red), Defender (blue), Goalkeeper (green), and Midfield (purple). Trend lines for each position group illustrate the general tendency in market value relative to playing time.

### Insight

Generally, players who accumulate more playing time tend to have a higher market value, as indicated by the upward sloping trend lines across all positions. However, the rate of increase in market value with minutes played appears to be steeper for attacking and midfield players compared to defenders and goalkeepers. This suggests that for outfield positions, consistent playing time is more strongly correlated with higher valuation. There are also notable outliers, particularly among attacking and midfield players, who command very high market values even with varying amounts of playing time, potentially reflecting exceptional talent or potential.

### Goals, Market Value & Position

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
# Load required libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyverse)
library(scales)

# Aggregate player appearances
player_stats <- appearances_top5 %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_goals = sum(goals, na.rm = TRUE),
    total_appearances = n(),
    total_minutes = sum(minutes_played, na.rm = TRUE),
    avg_minutes_per_appearance = total_minutes / total_appearances,
    .groups = "drop"
  )

player_values <- valuations_season %>%
  group_by(player_id) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE), .groups = "drop")

player_analysis <- player_stats %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(select(players, player_id, position), by = "player_id")

# Filter out goalkeepers from the player analysis
player_analysis <- player_analysis %>%
  filter(position != "Goalkeeper" & position != "Missing" & !is.na(position))  # Exclude goalkeepers

# Plot: Market Value vs. Goals
goal_plot <- ggplot(player_analysis, aes(x = total_goals, y = avg_market_value, color = position)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Market Value vs. Number of Goals",
    subtitle = "Top 5 European Leagues (2024–2025 Season)",
    x = "Total Goals",
    y = "Avg. Market Value (EUR)",
    color = "Position"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

print(goal_plot)

```


Figure 4: This scatterplot illustrates the correlation between total goals scored and average market value (in millions of Euros) for players in the top 5 European leagues during the 2024-2025 season. Each data point represents a player, distinguished by their position: Attack (red), Defender (blue), and Midfield (green). Trend lines for each position category highlight the general pattern in market value relative to the number of goals scored.

### Insight

This graph suggests a positive correlation between the number of goals scored and a player's market value, particularly seen in the upward trend lines for attacking and midfield positions. Goalkeepers were omitted due to their lack of goals. For attacking players, the market value tends to increase more significantly with each goal, showing that a higher premium is placed on better goal-scoring ability. Midfielders also show a positive trend, though less steep than attackers. On the other hand, defenders show a relatively flat trend line with fewer goals, indicating that their market value is less directly influenced by the number of goals they score. This suggests that factors other than goal-scoring, such as defensive skills, may contribute more to their valuation.

### Goal Distribution by Position

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
# Load required libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyverse)

# Filter for Top 5 European leagues
top5_leagues <- c("GB1", "FR1", "IT1", "ES1", "L1")

if ("competition_id" %in% colnames(appearances)) {
  appearances <- appearances %>%
    filter(competition_id %in% top5_leagues)
}

# Convert date column to Date format
if ("date" %in% colnames(appearances)) {
  appearances <- appearances %>%
    mutate(date = as.Date(date))
}

# Define season boundaries
season_start <- as.Date("2024-08-01")
season_end <- as.Date("2025-06-30")

# Summarize goal contributions per player
player_stats <- appearances %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_goals = sum(goals, na.rm = TRUE),
    total_assists = sum(assists, na.rm = TRUE),
    total_goal_contributions = total_goals + total_assists
  )

# Merge with player positions and remove goalkeepers
player_analysis <- player_stats %>%
  left_join(
    select(players, player_id, position),
    by = "player_id"
  ) %>%
  filter(position != "Goalkeeper")  # Exclude goalkeepers

# Generate box plot
box_plot <- ggplot(player_analysis, aes(x = position, y = total_goal_contributions, fill = position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 20)) +
  labs(
    title = "Distribution of Goal Contributions by Position",
    subtitle = "Top 5 European Leagues (2024–2025 Season)",
    x = "Position",
    y = "Total Goal Contributions"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 15, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

# Display the plot
print(box_plot)

```

Figure 5: This boxplot displays the distribution of total goal contributions across different player positions in the top 5 European leagues during the 2024-2025 season. The boxes represent the interquartile range (IQR), the horizontal line inside each box indicates the median, and the whiskers extend to show the variability outside the lower and upper quartiles. Outliers, if any, would be plotted as individual points beyond the whiskers.

### Insight

The distribution of goal contributions varies significantly by position. Attacking players exhibit the highest median number of goal contributions and the widest spread. This shows that while many contribute a moderate number of goals and assists, there is a strong tendency to achieve considerably higher totals. Midfielders show a lower median and a narrower IQR compared to attackers, meaning they have a more concentrated range of goal contributions. Defenders generally have the lowest median number of goal contributions, with a very tight IQR close to zero, which is expected given their primary role is not attacking. 

The presence of some outliers for defenders and midfielders also display instances where players in these positions have made surprisingly high attacking contributions.

### Top 10 Players by Market Value, Position, Goals, Appearances & Minutes Per Appearances 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}

# Load additional library
library(knitr)

# Aggregate appearances data by player
player_stats <- appearances %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_goals = sum(goals, na.rm = TRUE),
    total_appearances = n(),
    total_minutes = sum(minutes_played, na.rm = TRUE),
    avg_minutes_per_appearance = total_minutes / total_appearances
  )

# Get average market value per player
player_values <- valuations_season %>%
  group_by(player_id) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE))

# Combine player data
player_analysis <- player_stats %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(select(players, player_id, name, position), by = "player_id") %>%
  mutate(goal_contributions = total_goals)

# Select top 10 by market value
top10_players <- player_analysis %>%
  arrange(desc(avg_market_value)) %>%
  slice_head(n = 10) %>%
  mutate(
    avg_market_value_millions = round(avg_market_value / 1e6, 2)
  ) %>%
  select(
    name,
    position,
    total_goals,
    total_appearances,
    total_minutes,
    avg_minutes_per_appearance,
    avg_market_value_millions
  )

# Rename for table output
colnames(top10_players) <- c(
  "Player Name",
  "Position",
  "Goals",
  "Appearances",
  "Minutes Played",
  "Avg. Minutes per Appearance",
  "Avg. Market Value (EUR, Millions)"
)

# Display as markdown table
kable(top10_players, format = "markdown", align = "c", caption = "Top 10 Players by Average Market Value (2024–2025 Season)")

```

Figure 6: This table presents the top 10 players by average market value for the 2024-2025 season. It includes key performance indicators such as position, goals scored, total appearances, total minutes played, average minutes per appearance, and average market value in millions of Euros.

### Insight

The cumulative table highlights that attacking players generally command the highest market values, occupying six of the top ten spots. Vinicius Junior and Erling Haaland share the highest average market value at €200 million. Even though goals scored appear to be a significant factor in valuation (as seen with Haaland and Mbappé), other factors such as overall performance, and position also have a crucial role. However, strong midfielders also share high market values like the highly valued midfielders like Jude Bellingham and Jamal Musiala. Interestingly, Rodri, a midfielder with a high market value, has played significantly fewer minutes and scored no goals in the appearances listed, suggesting his valuation is likely based on his crucial role and impact beyond direct goal contributions.


### Market Value & Win Percentages

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Define top 5 European football leagues
top5_leagues <- c("GB1", "FR1", "IT1", "ES1", "L1")  # Premier League, Ligue 1, Serie A, La Liga, Bundesliga

# Define season date range
season_start <- as.Date("2024-08-01")
season_end <- as.Date("2025-06-30")

# Clean and filter appearances data
appearances_top5 <- appearances %>%
  filter(competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Clean and filter player valuations data
player_valuations_top5 <- player_valuations %>%
  filter(player_club_domestic_competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Prepare player data with standardized dates - convert from string format
players <- players %>%
  mutate(birth_date = as.Date(substr(date_of_birth, 1, 10)))  # Extract date part from datetime string

# Filter valuations for current season
valuations_season <- player_valuations_top5 %>%
  filter(between(date, season_start, season_end)) %>%
  left_join(players, by = "player_id", suffix = c("", "_player")) %>%
  mutate(age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25)

#cleaning player_valuations data
player_valuations_top5 <- player_valuations %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= season_start & date <= season_end, 
         player_club_domestic_competition_id %in% top5_leagues)

# filter player valuations to the correct leagues and date range
valuations_season <- player_valuations %>%
  mutate(date = as.Date(date)) %>%
  filter(player_club_domestic_competition_id %in% top5_leagues,
         date >= season_start & date <= season_end)

# calculate average market value per club during the season
club_market_values <- valuations_season %>%
  group_by(current_club_id) %>%
  summarise(
    avg_market_value = mean(market_value_in_eur, na.rm = TRUE),
    n_players = n()
  ) %>%
  ungroup()

# filter and prepare match results for the same season and leagues
games_season <- games %>%
  mutate(date = as.Date(date)) %>%
  filter(competition_id %in% top5_leagues,
         date >= season_start & date <= season_end)

# determine match winner 
games_with_results <- games_season %>%
  mutate(winner = case_when(
    home_club_goals > away_club_goals ~ home_club_id,
    away_club_goals > home_club_goals ~ away_club_id,
    TRUE ~ NA_integer_
  ))

# convert match data to team-level records
team_game_stats <- games_with_results %>%
  pivot_longer(cols = c(home_club_id, away_club_id),
               names_to = "home_away", values_to = "club_id") %>%
  group_by(club_id) %>%
  summarise(
    games_played = n(),
    wins = sum(club_id == winner, na.rm = TRUE),
    win_percentage = wins / games_played
  ) %>%
  ungroup()

# merge market value with win percentage and club names
club_performance <- team_game_stats %>%
  left_join(club_market_values, by = c("club_id" = "current_club_id")) %>%
  left_join(clubs, by = "club_id")  #assumes clubs has club_id and name

# plot market value vs win percentage with club labels
ggplot(club_performance, aes(x = win_percentage, y = avg_market_value)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_y_continuous(
    labels = function(x) paste0(round(x / 1e6), "M")
  ) +
  labs(
    title = "Do Teams With Higher Market Values Win More?",
    subtitle = "Top 5 European Leagues, 2024–25 Season",
    x = "Win Percentage",
    y = "Average Market Value (millions of eur)"
  ) +
  theme_minimal()
```
    
Figure 6 illustrates the relationship between a football club’s average market value (y-axis, in millions of euros) and its win percentage during the 2024–25 season (x-axis), across the top five European leagues. Each point represents a single club. The blue line represents a linear trend fitted to the data.

From the graph, we can observe a positive linear trend: as clubs' market values increase, their win percentages also tend to rise. The scatterplot shows a broad spread in both variables, but the upward slope of the regression line suggests that wealthier teams tend to perform better on average. Still, the data points are widely dispersed, indicating that market value alone does not perfectly predict win rate — several low-market-value clubs perform well, and a few high-value clubs underperform.

This visualization aligns with the well-documented assumption in football analytics that financial resources correlate with success, as wealthier clubs can afford more talented or experienced players. However, the wide dispersion hints at the influence of other factors that may enable less expensive teams to compete effectively or cause expensive squads to fall short.

When this figure is considered alongside Table 1 (Top Teams Table), patterns emerge suggesting that average player age, minutes played, and goal contribution may be mediators in this relationship. For example, younger teams with high contributions might be delivering high value per euro spent, signaling smart investment strategies.

This raises hypotheses worth exploring further:
- Does investment in specific positions better explain performance than raw value?
- Are teams with more balanced squad utilization (minutes played) more successful?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(gt)

# prepare players with birth_date
players_cleaned <- players %>%
  mutate(birth_date = as.Date(substr(date_of_birth, 1, 10))) %>%
  select(player_id, birth_date)

# join players to appearances to get age and contribution
appearances_with_stats <- appearances %>%
  mutate(date = as.Date(date)) %>%
  left_join(players_cleaned, by = "player_id") %>%
  mutate(
    age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25,
    goal_contribution = goals + assists
  )

# summarize per club
team_stats <- appearances_with_stats %>%
  group_by(player_club_id) %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    avg_minutes = mean(minutes_played, na.rm = TRUE),
    avg_goal_contribution = mean(goal_contribution, na.rm = TRUE),
    .groups = "drop"
  )

# merge with win percentage and club name
top_teams_table <- club_performance %>%
  select(club_id, win_percentage) %>%
  left_join(team_stats, by = c("club_id" = "player_club_id")) %>%
  left_join(clubs %>% select(club_id, name), by = "club_id") %>%
  select(name, win_percentage, avg_age, avg_minutes, avg_goal_contribution) %>%
  arrange(desc(win_percentage)) %>%
  slice_head(n = 15)

# view table
top_teams_table %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 2) %>%
  tab_header(
    title = "Top 15 Teams by Win Percentages and Player Metrics"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )
```

# Conclusion

This project examined how player attributes influence market value and team success in top European leagues. We found that attackers peak earlier and hold higher market values, while goal contributions and minutes played are strong indicators of valuation. Team-level data showed that clubs with higher average player values tend to win more. Overall, the analysis highlights how performance metrics and positional roles shape both individual worth and team outcomes in modern football.


### Code Appendix


```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Load required libraries
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(scales)
library(knitr)
library(gt)

# Load datasets
appearances <- read_csv("data/appearances.csv")
players <- read_csv("data/players.csv")
player_valuations <- read_csv("data/player_valuations.csv")
games <- read_csv("data/games.csv")
clubs <- read_csv("data/clubs.csv")

# Define top 5 European leagues and season dates
top5_leagues <- c("GB1", "FR1", "IT1", "ES1", "L1")
season_start <- as.Date("2024-08-01")
season_end <- as.Date("2025-06-30")

# Clean and filter datasets
appearances_top5 <- appearances %>%
  filter(competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

players <- players %>%
  mutate(birth_date = as.Date(substr(date_of_birth, 1, 10)))

player_valuations_top5 <- player_valuations %>%
  filter(player_club_domestic_competition_id %in% top5_leagues) %>%
  mutate(date = as.Date(date))

# Calculate player age and prepare valuations
valuations_season <- player_valuations_top5 %>%
  filter(date >= season_start & date <= season_end) %>%
  left_join(players %>% select(player_id, birth_date), by = "player_id") %>%
  mutate(age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25)

# Aggregate season stats
player_season_stats <- appearances_top5 %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_minutes = sum(minutes_played, na.rm = TRUE),
    matches_played = n(),
    avg_minutes_per_match = total_minutes / matches_played,
    .groups = "drop"
  )

# Merge most recent valuation and position
player_analysis <- valuations_season %>%
  group_by(player_id) %>%
  arrange(desc(date)) %>%
  slice(1) %>%
  ungroup() %>%
  left_join(player_season_stats, by = "player_id") %>%
  left_join(players %>% select(player_id, position), by = "player_id")

# Age binning
player_analysis <- player_analysis %>%
  mutate(
    age_bin = cut(age, breaks = seq(16, 40, by = 2), include.lowest = TRUE),
    age_rounded = floor(age)
  )

# Age vs Minutes scatter plot
ggplot(player_analysis, aes(x = age, y = total_minutes)) +
  geom_point(aes(color = position), alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, aes(color = position), alpha = 0.2) +
  facet_wrap(~ position, ncol = 2) +
  scale_color_manual(values = c("Attack" = "#FF5733", "Midfield" = "#33A1FD", "Defender" = "#9370DB", "Goalkeeper" = "#2ECC71")) +
  labs(title = "Age vs Minutes Played", x = "Age", y = "Total Minutes")

# Heatmap: Age bin × Position vs Market Value
heatmap_data <- player_analysis %>%
  filter(!is.na(market_value_in_eur) & !is.na(position) & !is.na(age_bin)) %>%
  group_by(age_bin, position) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE), .groups = "drop") %>%
  mutate(avg_market_value_millions = avg_market_value / 1e6)

all_combinations <- expand.grid(age_bin = levels(player_analysis$age_bin), position = unique(player_analysis$position))

heatmap_data <- all_combinations %>%
  left_join(heatmap_data, by = c("age_bin", "position")) %>%
  mutate(label = ifelse(is.na(avg_market_value_millions), "N/A", sprintf("%.0f€", avg_market_value_millions)))

heatmap_data$position <- factor(heatmap_data$position, levels = c("Goalkeeper", "Defender", "Midfield", "Attack", "Unknown"))

ggplot(heatmap_data, aes(x = age_bin, y = position, fill = avg_market_value_millions)) +
  geom_tile() +
  geom_text(aes(label = label), color = "white") +
  scale_fill_gradient(low = "blue", high = "orange", na.value = "grey50", name = "Avg. Value (M€)")

# Market Value vs Minutes Played
player_values <- valuations_season %>%
  group_by(player_id) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE))

player_stats <- appearances_top5 %>%
  filter(date >= season_start & date <= season_end) %>%
  group_by(player_id) %>%
  summarise(
    total_minutes = sum(minutes_played, na.rm = TRUE),
    .groups = "drop"
  )

plot_data <- player_stats %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(players %>% select(player_id, position), by = "player_id")

ggplot(plot_data, aes(x = total_minutes, y = avg_market_value, color = position)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M"))

# Market Value vs Goals
goal_data <- appearances_top5 %>%
  group_by(player_id) %>%
  summarise(total_goals = sum(goals, na.rm = TRUE)) %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(players %>% select(player_id, position), by = "player_id") %>%
  filter(position != "Goalkeeper")

ggplot(goal_data, aes(x = total_goals, y = avg_market_value, color = position)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE)

# Boxplot: Goal Contributions by Position
goal_contributions <- appearances_top5 %>%
  group_by(player_id) %>%
  summarise(
    total_goals = sum(goals, na.rm = TRUE),
    total_assists = sum(assists, na.rm = TRUE),
    total_goal_contributions = total_goals + total_assists
  ) %>%
  left_join(players %>% select(player_id, position), by = "player_id") %>%
  filter(position != "Goalkeeper")

ggplot(goal_contributions, aes(x = position, y = total_goal_contributions, fill = position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA)

# Table: Top 10 Players by Market Value
top10 <- player_stats %>%
  inner_join(player_values, by = "player_id") %>%
  left_join(players %>% select(player_id, name, position), by = "player_id") %>%
  mutate(goal_contributions = total_goals) %>%
  arrange(desc(avg_market_value)) %>%
  slice(1:10)

kable(top10, format = "markdown")

# Club-Level Market Value vs Win %
games_with_results <- games %>%
  mutate(date = as.Date(date)) %>%
  filter(competition_id %in% top5_leagues, date >= season_start & date <= season_end) %>%
  mutate(winner = case_when(
    home_club_goals > away_club_goals ~ home_club_id,
    away_club_goals > home_club_goals ~ away_club_id,
    TRUE ~ NA_integer_
  ))

team_results <- games_with_results %>%
  pivot_longer(cols = c(home_club_id, away_club_id), names_to = "home_away", values_to = "club_id") %>%
  group_by(club_id) %>%
  summarise(games_played = n(), wins = sum(club_id == winner, na.rm = TRUE), win_percentage = wins / games_played)

club_values <- valuations_season %>%
  group_by(current_club_id) %>%
  summarise(avg_market_value = mean(market_value_in_eur, na.rm = TRUE))

ggplot(team_results %>% left_join(club_values, by = c("club_id" = "current_club_id")),
       aes(x = win_percentage, y = avg_market_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

# Top 15 Clubs Table
player_club_stats <- appearances %>%
  mutate(date = as.Date(date)) %>%
  left_join(players %>% select(player_id, birth_date), by = "player_id") %>%
  mutate(age = as.numeric(difftime(date, birth_date, units = "days")) / 365.25,
         goal_contribution = goals + assists) %>%
  group_by(player_club_id) %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    avg_minutes = mean(minutes_played, na.rm = TRUE),
    avg_goal_contribution = mean(goal_contribution, na.rm = TRUE)
  )

top_club_table <- team_results %>%
  left_join(player_club_stats, by = c("club_id" = "player_club_id")) %>%
  left_join(clubs %>% select(club_id, name), by = "club_id") %>%
  arrange(desc(win_percentage)) %>%
  slice(1:15) %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 2) %>%
  tab_header(title = "Top 15 Teams by Win Percentages and Player Metrics") %>%
  cols_align(align = "center", columns = everything())
```
